---
title: "Simulate data for CCW project"
bibliography: references.bib
---

```{r }
#| label: setup
#| echo: false
#| message: false
#| 
library(here)
library(ggdag)
library(corrplot)

source(here('scripts', 'setup.R'), echo=F)
# Set seed for reproducibility
set.seed(42)
```

# Causal Directed Acyclic Graph  

We assume a generic treatment, A, has no causal effect on generic outcome, Y. However, confounding is present through another random variable. The DAG includes `age`, `X` a confounder, sex defined through a binary variable `F` for female. X is both a baseline variable, and a time-varying confounder. Below we assign effects of these variables that will lead to identification of a causal effect between A and Y, if no controlling for the confounding is done. 

```{r }
#| label: gg-dag
#| fig-cap: "Figure 1. Causal DAGs of treatment and outcome"
#| fig-subcap: 
#|  - 'A causes Y'
#|  - 'A not a cause of Y'
#| layout-ncol: 2
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: false
#| 
dag_ay <- dagify(
  Y ~ Age,
  Y ~ F,
  Y ~ A,
  A ~ X,
  Y ~ X,
  labels = c(
    "Age" = "Age",
    "F" = "Female",
    "A" = "A",
    "Y" = "Y",
    "X" = "X"
  ),
  exposure = "A",
  outcome = "Y"
)

ggdag(dag_ay, text = FALSE, use_labels = "label") +
  theme_dag()

dag_xy <- dagify(
  Y ~ Age,
  Y ~ F,
  A ~ X,
  Y ~ X,
  labels = c(
    "Age" = "Age",
    "F" = "Female",
    "A" = "A",
    "Y" = "Y",
    "X" = "X"
  ),
  exposure = "A",
  outcome = "Y"
)

ggdag(dag_xy, text = FALSE, use_labels = "label") +
  theme_dag()
```

Two basic scenarios are mapped out. 1) Where treatment A causes Y but a confounder X is present, 2) Treatment A doesn't cause Y, but confounder is present. In a naive analysis of either (no adjustment), the estimate will be biased because treatment and outcome are d-connected.  

```{r }
#| label: gg-dag-dsep-2
#| fig-cap: "Figure 2. D-separated treatment and outcome after adjustment"
#| fig-cap-location: top
#| fig-subcap: 
#|  - 'A causes Y'
#|  - 'A not a cause of Y'
#| layout-ncol: 2
#| message: false
#| warning: false
#| echo: false


ggdag_dseparated(dag_ay, text = FALSE, use_labels = "label",
                 controlling_for = c('X')) +
  theme_dag()

ggdag_dseparated(dag_xy, text = FALSE, use_labels = "label",
                 controlling_for = c('X')) +
  theme_dag()

```

Adjustment for `X` should lead to identification of a *NULL* effect of treatment on outcome in scenario 2 but not 1.

# Simulation Parameters {#sec-simparams} 

```{r }
#| label: sim-startdta
#| message: false
#| warning: false
#| 

set.seed(100)
n = 2000L # <1>
fup = 61 # <1>

df = tibble(id = 1:n, # <2>
            age = round(rnorm(n, mean = 75, sd = 10), 1), # <2> 
            female = sample(c(1, 0), size = n, replace = TRUE, prob = c(0.66, 0.34)), # <2>
            fup = fup, # <2> 
            X = rbinom(n, 1, 0.05), # <2>
            ) # <2> 
  
setDT(df) 
d_panel = df[rep(seq(.N), fup)] # <3> 
d_panel[, exit := (seq_len(.N)), by = list(id)] # <3>
d_panel[, enter := exit-1] # <3> 
d_panel[, time := seq_len(.N), by = list(id)] # <3>
```

1. Specify `r n` individuals with `r fup` observation periods of follow-up. 
2. Specify covariates, `age`, `female` and `X`.
3. Expand dataset, each observation period is an interval of length = 1.

For baseline `X`, further include `X_t` which varies across time. A change in X_t is triggered by intermediate variable `X_shift` which is a random binomial draw (meant to represent a triggering event for a change in X). So for each person, there is a baseline X, then a random timepoint in follow-up where X changes. 

```{r }
#| label: sim-tvarydata
#| message: false
#| warning: false

d_panel[, X_t := rbinom(.N, 1, 0.03)]
```

```{r }
#| label: gg-tvary
#| fig-cap: "Figure 4. Distribution of Time-varying Confounder"
#| fig-cap-location: top
#| echo: false
#| eval: false
d_panel %>%
  ggplot(aes(x = cut_width(time, 12), y = X_t)) +
  geom_violin( scale = "width", fill = cbbPalette[1], alpha = 0.5) +
  stat_summary(fun = "mean",
               geom = "point",
               color = cbbPalette[2], size=2) +
  geom_hline(aes(yintercept=0),linewidth=1.1, linetype=3) +
  scale_x_discrete(labels = seq(0, 90, 12)) +
  labs(x = "Follow-up", y = "Time-varying X", ) +
  theme_bw()
```

Time to treatment and outcome is simulated in a longitudinal dataset, to allow for time-varying effect of X. Then the dataset is modified so that persons follow-up ends at the time of outcome:  

```{r }
#| label: sim-treatoutc
#| message: false
#| warning: false
  set.seed(101)
  
d_panel_2 = d_panel %>% 
    # Treat event # <1>
    mutate(log_odds = -4.1 + 0.0001*time + -0.00001*(time^2) + 1.2*X + 1.5*X_t,
           p = exp(log_odds) / (1 + exp(log_odds)),
           treat = rbinom(n(), size = 1, prob = p),
           treat = if_else(treat==1, 1, NA_integer_)
    ) %>%
    group_by(id) %>%
      fill(treat, .direction = 'down') %>%
    ungroup %>%
    mutate(treat = coalesce(treat, 0L)) # <1>
  
# Dataset with no A -> Y effect
  d_xy = d_panel_2 %>%
    mutate(model = 'xy') %>%
    # Outcome event # <2>
    mutate(log_odds = -5 + 0.007*time + 0.0002*(time^2) + 0.6*X + 0.9*X_t + 
                                  -0.2*female + 0.07*age + -0.001*(age^2),
           p = exp(log_odds) / (1 + exp(log_odds)),
           outcome = rbinom(n(), size = 1, prob = p),
           outcome = if_else(outcome==1, 1, NA_integer_)
    ) %>%
    group_by(id) %>%
      fill(outcome, .direction = 'down') %>% 
    ungroup %>%
    mutate(outcome = coalesce(outcome, 0L)) %>% # <2>
    group_by(id) %>% # <3>
      mutate(t_outcome = coalesce(min(time[outcome==1]), Inf),
             t_treat   = coalesce(min(time[treat==1]), Inf)) %>%
    ungroup %>%
    dplyr::filter(outcome  == 0 | time == t_outcome) %>%
    dplyr::filter(time <= 60) %>%
    mutate(t_treat = if_else(t_outcome < t_treat, Inf, t_treat),
           t_treat = if_else(60 < t_treat, Inf, t_treat))  # <3>
  
# Dataset with A -> Y effect
  d_ay = d_panel_2 %>%
    mutate(model = 'ay') %>%
    # Outcome event # 
    mutate(log_odds = -5 + 0.007*time + 0.0002*(time^2) + 0.6*X + 0.9*X_t + 
                                  -0.2*female + 0.07*age + -0.001*(age^2) + treat*-1.2, # <4>
           p = exp(log_odds) / (1 + exp(log_odds)),
           outcome = rbinom(n(), size = 1, prob = p),
           outcome = if_else(outcome==1, 1, NA_integer_)
    ) %>%
    group_by(id) %>%
      fill(outcome, .direction = 'down') %>% 
    ungroup %>%
    mutate(outcome = coalesce(outcome, 0L)) %>% # 
    group_by(id) %>% # <3>
      mutate(t_outcome = coalesce(min(time[outcome==1]), Inf),
             t_treat   = coalesce(min(time[treat==1]), Inf)) %>%
    ungroup %>%
    dplyr::filter(outcome  == 0 | time == t_outcome) %>%
    dplyr::filter(time <= 60) %>%
    mutate(t_treat = if_else(t_outcome < t_treat, Inf, t_treat),
           t_treat = if_else(60 < t_treat, Inf, t_treat))  
  
  d_panel_3 = bind_rows(d_ay, d_xy)
```
1. Simulation of treatment event times. Treatment is only a function of `time`, `X`, and `X_t`.
2. Simulation of outcome event times. Outcome is a function of `time`, `X`, and `X_t`, `F`, `Age`.
3. Modify dataset so that follow-up times coincide with new treatment and outcome variables. 
4. Additional term for treatment for modeling a causal effect.

### Simulated treatment times  

```{r }
#| label: summ-ttimes
#| message: false
#| warning: false
#| echo: false

d_panel_3 %>%
  group_by(model) %>%
  summarize(min = min(t_treat),
            q05 = quantile(t_treat, 0.05),
            q25 = quantile(t_treat, 0.25),
            q50 = quantile(t_treat, 0.50),
            q75 = quantile(t_treat, 0.75),
            q95 = quantile(t_treat, 0.95),
            max = max(t_treat)
            ) %>%
  t() %>%
  kbl(caption = "Time to treatment by causal model") %>%
  kable_styling()
```

### Simulated outcome times  

```{r }
#| label: summ-otimes
#| message: false
#| warning: false
#| echo: false
d_panel_3 %>%
  group_by(model) %>%
  summarize(min = min(t_outcome),
            q05 = quantile(t_outcome, 0.05),
            q25 = quantile(t_outcome, 0.25),
            q50 = quantile(t_outcome, 0.50),
            q75 = quantile(t_outcome, 0.75),
            q95 = quantile(t_outcome, 0.95),
            max = max(t_outcome)
            ) %>%
  t() %>%
  kbl(caption = "Time to treatment by causal model") %>%
  kable_styling()
```

```{r }
#| label: gg-treatoutc
#| fig-cap: "Figure 5. Treatment and Outcome Incidence Across Follow-up"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: false
gg_treat = d_panel_3 %>%
  group_by(model, time) %>%
  summarize(treat_prob = mean(treat)) %>%
  ggplot(aes(x = time, y = treat_prob)) +
  geom_line(aes(color = model)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = 'Follow-up', y = 'Proportion treated', title = 'Treatment') +
  theme_bw()

gg_outc = d_panel_3 %>%
  group_by(model, time) %>%
    summarize(outcome = sum(outcome)) %>%
  ggplot(aes(x = time, y = outcome)) +
  geom_point(aes(color = model)) +
  labs(x = 'Follow-up', y = 'Events', title = 'Outcome') +
  theme_bw()

ggarrange(gg_treat, gg_outc, ncol = 1)
```

```{r }
#| label: gg-corplot
#| fig-cap: "Figure 6. Correlation plot of covariates, treatment, outcome (A->Y model)"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: false
corrplot(cor(d_panel_3[d_panel_3$model =='xy', c('age', 'female', 'X', 'X_t', 'treat', 'outcome')]),
         method = 'number', diag=F, type='upper', bg = 'black')
```

# Summary of survival dataset {#sec-summdta}

```{r }
#| label: glimpse-survdta
#| echo: false

glimpse(d_panel_3)

d_sv = select(d_panel_3, model, id, age, female, fup, X, X_t, treat, enter, exit, time, outcome, t_outcome, t_treat)

saveRDS(d_sv, here('dta', 'dta_cohort.Rds'))
```

# Cloning procedure {#sec-cloning}

The cloning procedure is very project specific, tied to how the treatment strategy is defined so it is hard to give general recommendations here. For this tutorial, we describe an arbitrary window in which treatment is expected to initiate within 12 time periods and outline strategies around this:

**Treatment Strategies: **  

1) Do not ever receive treatment  

2) Initiate treatment within 12 time periods (days, weeks etc.) and if not then treatment will initiate on week 12  

::: callout-note
The grace window is a funny concept when you first consider it. This does not reflect any trial I have ever heard of actually being done but may identify an interesting or important effect. It is essentially outlying a treatment "protocol" and saying what if everyone adhered to this protocol counterfactual to what was observed or some other intervention (i.e. no one initiates treatment).
:::

To clone, you make two copies of the data, and make a new `artifical censoring` variable, which is a censoring time for the period when clones observed data are no longer consistent with their assigned strategy.  

```{r }
#| label: clone-dta
#| message: false
#| warning: false
data_cloned = bind_rows( # <1>
                     d_panel_3 %>% # <1>
                       mutate(assign = 0, # <1>
                              censor = if_else(t_treat <= time, 1L, 0L), # <1>
                              event = if_else(censor==0 & t_outcome<=time, 1L, 0L), # <1>
                       ), # <1>
                     d_panel_3 %>% # <2>
                       mutate(assign = 1, # <2>
                              censor = if_else(t_treat > 12 & time>=12, 1L, 0L), # <2>
                              event  = if_else(censor==0 & t_outcome<=time, 1L, 0L)
                              ) # <2>
                       ) %>% # <2>
  arrange(model, id, assign, time) %>%
  group_by(model, id, assign) %>%
    mutate(t_censor = min(time[censor==1])) %>%
    dplyr::filter(censor == 0 | time == min(time[censor==1])) %>% # <3>
  ungroup %>%
  select(model, id, time, t_outcome, t_treat, t_censor, censor, event, assign, enter, exit, everything(), -fup, -log_odds, -p)
```

1. Clones assigned to strategy 1 (No treatment)
2. Clones assigned to strategy 2 (grace window for treatment)
3. Update failure times and events counting for artificial censoring

::: callout-note
Note how we do not specify death or some other competing  event, competing events complicate things but are common in applied work. I keep it simple here. 
:::

# Summary of simulated, cloned data-set  

```{r }
#| label: summ-lone-dta
#| message: false
#| warning: false
#| echo: false
data_cloned %>%
  group_by(model, assign) %>%
  summarize(`Unique persons` = n_distinct(id),
            `Median (Follow-up)` = median(last(time)),
            `Total treat (ever)` = sum(n_distinct(id[t_treat<Inf])),
            `Total treat (t<=12)` = sum(n_distinct(id[t_treat<=12])),
            `Total events` = sum(event),
            `Total censored` = sum(censor)) %>%
  t() %>%
  kable() %>%
  kable_styling()
```

A person-level version of the data is also created (excluded time-varying X) for comparison:  

```{r }
#| label: sv-personlevel
#| message: false
#| warning: false
#| 
data_cloned_p = data_cloned %>%
  group_by(model, id, assign) %>%
    mutate(t_censor = min(time[censor==1])) %>%
  ungroup %>%
  mutate(time = pmin(t_censor, t_outcome, 60),
         event = case_when(
           time==t_censor ~ 0, 
           time==t_outcome ~ 1,
           T ~ 0)) %>%
  select(model, id, assign, t_treat, t_outcome, t_censor, time, event, X, female, age) %>%
  distinct()
```

```{r }
#| label: sv-panel
#| message: false
#| warning: false
#| echo: false
saveRDS(data_cloned_p, here('dta', 'dta_cloned_person.Rds'))

saveRDS(data_cloned, here('dta', 'dta_cloned_panel.Rds'))
```