---
title: "Presentation"
subtitle: "Tables and Figures for Manuscript"
bibliography: references.bib
---

```{r }
#| label: setup
#| echo: false
#| message: false
#| 
library(here)
library(purrr)

source(here('scripts', 'setup.R'), echo=F)
# Set seed for reproducibility
set.seed(42)

dta_c_panel = readRDS(here('dta', 'dta_cloned_panel.Rds')) %>%
  dplyr::filter(model == 'ay')
```

# Description of cohort, “Table 1”

In both randomized trials and observational studies comparing interventions/treatments there is typically a presentation of the overall cohort, or comparison of those receiving treatment versus alternatives. The table will present basic demographics and some additional characteristics of interest (confounders, important predictors/causes of outcome etc.). I generally do not present inferential statistics between groups in this table (i.e. no p-values or confidence intervals) but will provide a standardized mean difference.

Some special considerations are needed when presenting target trial emulations.

## Cloning & Grace Period

The cloning presents an interesting problem for a “Table 1” because the groups are identical at baseline. Artificial censoring may change the distribution of covariates over time if censoring is correlated with these covariates. It probably makes sense for most projects using a clone-sensor-weight approach to present a “baseline” cohort column, and then columns describing the intervention groups at a key follow-up time point. The end of the grace period is a good standard choice. It makes sense to present the time-invariant covariates as well as time-varying covariates.

Assume a simple example where the design of the study includes a baseline period, "time zero" and two treatment strategies, with a grace period or interval where one strategy allows some time for treatment to occur.There being two main periods: 1) Baseline time zero, 2) End of the grace period.

Here is one approach:

1.  Present the baseline characteristics for everyone (prior to cloning) at time zero.
2.  Present the characteristics for both treatment groups at the end of the grace period.

This would lead to a "Table 1" with three columns.

## Table 1

Many R packages will construct a "Table 1" for users. I manually code below for transparency. 

```{r }
#| label: tab-1
#| warning: FALSE

# Many packages can do this for you, I write out manually for transparency

d_cbase = dta_c_panel %>%
  # Take one obs for the baseline column
  dplyr::filter(time==1 & assign==1 & censor==0) %>%
  summarize(N = prettyNum(n(), big.mark=','),
            X = paste0(round(mean(X), 2), ' (', round(sd(X), 2), ')'),
            X_t = paste0(round(mean(X_t), 2), ' (', round(sd(X_t), 2), ')'),
            `Age, years (SD)`  = paste0(floor(mean(age)), ' (', floor(sd(age)), ')'),
            `Sex, % Female` = paste0(prettyNum(sum(female==1), big.mark = ','), 
                                        ' (', round(mean(female)*100,1), ')')) %>%
  t()

d_c0 = dta_c_panel %>%
  # End of grace period
  dplyr::filter(time==12 & assign==0 & censor==0) %>%
  summarize(N = prettyNum(n(), big.mark=','),
            X = paste0(round(mean(X), 2), ' (', round(sd(X), 2), ')'),
            X_t = paste0(round(mean(X_t), 2), ' (', round(sd(X_t), 2), ')'),
            `Age, years (SD)`  = paste0(floor(mean(age)), ' (', floor(sd(age)), ')'),
            `Gender, % Female` = paste0(prettyNum(sum(female==1), big.mark = ','), 
                                        ' (', round(mean(female)*100,1), ')')) %>%
  t()
  
d_c1 = dta_c_panel %>%
  # End of grace period
  dplyr::filter(time==12 & assign==1 & censor==0) %>%
  summarize(N = prettyNum(n(), big.mark=','),
            X = paste0(round(mean(X), 2), ' (', round(sd(X), 2), ')'),
            X_t = paste0(round(mean(X_t), 2), ' (', round(sd(X_t), 2), ')'),
            `Age, years (SD)`  = paste0(floor(mean(age)), ' (', floor(sd(age)), ')'),
            `Sex, % Female` = paste0(prettyNum(sum(female==1), big.mark = ','), 
                                        ' (', round(mean(female)*100,1), ')')) %>%
  t()

d_tab1 = bind_cols(Variables = c('N', 'X, mean (SD)', 'Xt, mean (SD)', 'Age, years (SD)', 'Sex, female'),
          `Baseline` = d_cbase, `Assign=0` = d_c0, `Assign=1` = d_c1) 
```

```{r }
#| label: tab-1-kable
#| tbl-cap-location: top
#| tbl-cap: "Table 1. Example of presentation of characteristics of a CCW Cohort"
#| echo: false
d_tab1 %>%
  kable(, align = 'c') %>%
  kable_styling() %>%
  add_header_above(c(" " = 2, "Uncensored at end of grace period" = 2))
```

## Presentation of weighted differences

Most researchers would provide the naive differences between groups, and then present some adjusted difference. Most commonly the standardized mean differences (aSMD) or a P-value from a univariate test (not recommended).

So you could add one column for the SMD to the "Table 1", but there are also other approaches such as reporting Mahalanobis distance.

### Computation of distances

```{r }
#| label: tab-1-smd
#| warning: FALSE
#| 
d_smd = dta_c_panel %>%
  dplyr::filter(time==12 & censor==0) %>%
  summarize(`X, mean (SD)` = (mean(X[assign==1]) - mean(X[assign==0])) / sd(X),
            `Xt, mean (SD)` = (mean(X_t[assign==1]) - mean(X_t[assign==0])) / sd(X_t),
            `Age, years (SD)`  = (mean(age[assign==1]) - mean(age[assign==0])) / sd(age),
            `Sex, female` = (mean(female[assign==1]) - mean(female[assign==0])) / sd(female)
         ) %>%
  t() %>%
  as_tibble(rownames = 'Variable') %>%
  rename(`aSMD` = V1)

d_tab1 = left_join(d_tab1, d_smd, by=c('Variables' = 'Variable'))
```

```{r }
#| label: tab-1-mahdist
#| warning: FALSE
#| 

# Compute pooled covariance matrix
cov_pooled <- (var(dta_c_panel[dta_c_panel$assign==1, c('X', 'X_t', 'age', 'female')]) + 
                 var(dta_c_panel[dta_c_panel$assign==0, c('X', 'X_t', 'age', 'female')])) / 2

# Compute Mahalanobis distance
mahal_dist <- mahalanobis(colMeans(dta_c_panel[dta_c_panel$assign==1, 
                                               c('X', 'X_t', 'age', 'female')]), 
                          center = colMeans(dta_c_panel[dta_c_panel$assign==0, 
                                                        c('X', 'X_t', 'age', 'female')]), 
                          cov = cov_pooled)
```

### Table 1 with differences

```{r }
#| label: tab-1-smd-kable
#| tbl-cap-location: top
#| tbl-cap: "Table 1 + aSMD (unweighted)"
#| echo: false
#| 
  
kable(d_tab1, align = 'c', digits=3) %>%
  kable_styling() %>%
  add_header_above(c(" " = 2, "Uncensored at end of grace period" = 3)) 
```
Note: `r paste0("Mahalanobis distance between groups, ", 'Naive: ', round(mahal_dist, 3))`

### Table 1 with weighted differences

::: callout-note
See [Estimators](ccw_02_est.qmd) for full description of weight estimation
:::

```{r }
#| label: est-wts
#| tbl-cap-location: top
#| tbl-cap: "Table 1 + aSMD (unweighted)"
#| echo: false
#| warning: false
#| 
d_glm_wt = glm(treat ~ poly(time, 2, raw=T) + X + X_t + age + poly(age, 2) + female, # <1>
                 data=dta_c_panel[dta_c_panel$assign==0, ], family=binomial()) # <1>
  
dta_c_panel$pr_treat = predict(d_glm_wt, newdata = dta_c_panel, type='response') # <2>
  
setDT(dta_c_panel)
  
dta_c_panel[, ipw := fcase(
    assign==0 & censor==0, 1 / (1-pr_treat), # <1>
    assign==0 & censor==1, 0, # <2>
    assign==1 & censor==1, 0, # <2>
    assign==1 & time < 12, 1, # <3>
    assign==1 & time == 12  & t_treat  < 12, 1, # <4> 
    assign==1 & time == 12  & t_treat  ==12 & censor==0, 1 / (pr_treat), # <5> 
    assign==1 & time > 12, 1 # <6>
  )]
  
dta_c_panel[, marg_c0  := 1-mean(censor), by = list(assign, time)] # <1>

dta_c_panel[, marg_ipw := fcase(
    assign==0 & censor==0, marg_c0 / (1-pr_treat), # <2>
    assign==0 & censor==1, 0,  
    assign==1 & censor==1, 0,  
    assign==1 & time < 12, 1, 
    assign==1 & time == 12  & t_treat  < 12, 1,  
    assign==1 & time == 12  & t_treat  ==12 & censor==0, marg_c0 / (pr_treat), # <2> 
    assign==1 & time > 12, 1 
  )]

dta_c_panel[, `:=`(ipw = cumprod(ipw), 
                     marg_ipw = cumprod(marg_ipw)), 
               by=list(id, assign)] 
```

```{r }
#| label: tab1-smd-wts
#| tbl-cap-location: top
#| tbl-cap: "Table 1 + aSMD (weighted)"
#| echo: false
#| 
d_smd_wt = dta_c_panel %>%
  # End of grace period
  dplyr::filter(time==12 & censor==0) %>%
  summarize(`X, mean (SD)` = (weighted.mean(X[assign==1], w = ipw[assign==1]) - 
                     weighted.mean(X[assign==0], w = ipw[assign==0])) / sd(X),
            `Xt, mean (SD)`  = (weighted.mean(X_t[assign==1], w = ipw[assign==1]) - 
                      weighted.mean(X_t[assign==0], w = ipw[assign==0])) / sd(X_t),
            `Age, years (SD)` = (weighted.mean(age[assign==1], w = ipw[assign==1]) - 
                               weighted.mean(age[assign==0], w = ipw[assign==0])) / sd(age),
            `Sex, female` = (weighted.mean(female[assign==1], w = ipw[assign==1]) - 
                                  weighted.mean(female[assign==0], w = ipw[assign==0])) / sd(female)
         ) %>%
  t() %>%
  as_tibble(rownames = 'Variable') %>%
  rename(`IPW aSMD` = V1)

# Compute pooled covariance matrix
cov_pooled <- (var(dta_c_panel[dta_c_panel$assign==1, c('X', 'X_t', 'age', 'female')]) + 
                 var(dta_c_panel[dta_c_panel$assign==0, c('X', 'X_t', 'age', 'female')])) / 2

# Compute Mahalanobis distance
grp_1_mn = dta_c_panel %>%
  # End of grace period
  dplyr::filter(time==12 & censor==0 & assign==1) %>%
  summarize(X = weighted.mean(X, w = ipw),
            X_t = weighted.mean(X_t, w = ipw),
            age = weighted.mean(age, w = ipw),
            female = weighted.mean(female, w = ipw))

grp_0_mn = dta_c_panel %>%
  # End of grace period
  dplyr::filter(time==12 & censor==0 & assign==0) %>%
  summarize(X = weighted.mean(X, w = ipw),
            X_t = weighted.mean(X_t, w = ipw),
            age = weighted.mean(age, w = ipw),
            female = weighted.mean(female, w = ipw))

wt_mahal_dist <- mahalanobis(unlist(grp_1_mn), 
                          center = unlist(grp_0_mn), 
                          cov = cov_pooled)

d_tab1 = left_join(d_tab1, d_smd_wt, by = c('Variables' = 'Variable')) 
```

```{r }
#| label: tab-1-smdwtd-kable
#| tbl-cap-location: top
#| tbl-cap: "Table 1 + aSMD (weighted)"
#| echo: false
#| 
kable(d_tab1, align = 'c', digits=3) %>%
  kable_styling() %>%
  add_header_above(c(" " = 2, "Uncensored at end of grace period" = 4)) 
```

Note. `r paste0("Mahalanobis distance between groups, ", 'Naive: ', round(mahal_dist, 3), ', IPW: ', round(wt_mahal_dist, 3))`

## Figure of differences across time  

These differences can also be presented in a modified "Love" plot, with panels for each period of follow-up. It doesn't really make sense to do one at baseline because the differences will be small/none. But we can pick a few follow-up periods to evaluate:  

```{r }
#| label: mk-funcs
#| message: false
#| warning: false
#| echo: false
f_cpt_dist_wt = function(dta) {
  
  d_dist = dta %>%
  summarize(`X, mean (SD)` = (weighted.mean(X[assign==1], w = ipw[assign==1]) - 
                     weighted.mean(X[assign==0], w = ipw[assign==0])) / sd(X),
            `Xt, mean (SD)`  = (weighted.mean(X_t[assign==1], w = ipw[assign==1]) - 
                      weighted.mean(X_t[assign==0], w = ipw[assign==0])) / sd(X_t),
            `Age, years (SD)` = (weighted.mean(age[assign==1], w = ipw[assign==1]) - 
                               weighted.mean(age[assign==0], w = ipw[assign==0])) / sd(age),
            `Sex, female` = (weighted.mean(female[assign==1], w = ipw[assign==1]) - 
                                  weighted.mean(female[assign==0], w = ipw[assign==0])) / sd(female)
         ) %>%
  t() %>%
  as_tibble(rownames = 'Variable') %>%
  rename(`SMD` = V1)

  cov_pooled <- (var(dta[dta$assign==1, c('X', 'X_t', 'age', 'female')]) + 
                   var(dta[dta$assign==0, c('X', 'X_t', 'age', 'female')])) / 2

  grp_1_mn = dta %>%
    dplyr::filter(assign==1) %>%
    summarize(X = weighted.mean(X, w = ipw),
              X_t = weighted.mean(X_t, w = ipw),
              age = weighted.mean(age, w = ipw),
              female = weighted.mean(female, w = ipw))

  grp_0_mn = dta %>%
    dplyr::filter(assign==0) %>%
    summarize(X = weighted.mean(X, w = ipw),
              X_t = weighted.mean(X_t, w = ipw),
              age = weighted.mean(age, w = ipw),
              female = weighted.mean(female, w = ipw))

  wt_mahal_dist <- mahalanobis(unlist(grp_1_mn), 
                            center = unlist(grp_0_mn), 
                            cov = cov_pooled)
  
  bind_rows(d_dist, 
                     tibble(Variable = 'M-dist', `SMD` = wt_mahal_dist))
  
}

f_cpt_dist = function(dta) {
  
  d_dist = dta %>%
  summarize(`X, mean (SD)` = (mean(X[assign==1]) - mean(X[assign==0])) / sd(X),
            `Xt, mean (SD)` = (mean(X_t[assign==1]) - mean(X_t[assign==0])) / sd(X_t),
            `Age, years (SD)`  = (mean(age[assign==1]) - mean(age[assign==0])) / sd(age),
            `Gender, female` = (mean(female[assign==1]) - mean(female[assign==0])) / sd(female)
         ) %>%
  t() %>%
  as_tibble(rownames = 'Variable') %>%
  rename(`SMD` = V1)
    
  cov_pooled <- (var(dta[dta$assign==1, c('X', 'X_t', 'age', 'female')]) + 
                     var(dta[dta$assign==0, c('X', 'X_t', 'age', 'female')])) / 2

  mahal_dist <- mahalanobis(colMeans(dta[dta$assign==1, 
                                                 c('X', 'X_t', 'age', 'female')]), 
                            center = colMeans(dta[dta$assign==0, 
                                                          c('X', 'X_t', 'age', 'female')]), 
                            cov = cov_pooled)
  
  bind_rows(d_dist, tibble(Variable = 'M-dist', `SMD` = mahal_dist)) 

}
```

```{r }
#| label: fig-1-smdwtd
#| fig-cap-location: top
#| fig-cap: "Baseline and Time-varying Differences by Treatment Group"
#| message: false
#| warning: false
# aSMD Figure ----
  i_SMD_grps = c(12, 24, 60)

  d_time = dta_c_panel %>%
    dplyr::filter(time %in% c(i_SMD_grps) & censor==0) %>%
    nest(.by = time) 
  
  d_time$SMD_naive = map(d_time$data, ~f_cpt_dist(.))
  d_time$SMD_ipw = map(d_time$data, ~f_cpt_dist_wt(.))
  
  gg_bal = select(d_time, time, SMD_naive, SMD_ipw) %>%
    unnest() %>%
    select(-Variable1) %>%
    rename(`Naive` = SMD,
           `IPW` = SMD1) %>%
      pivot_longer(cols = c('Naive', 'IPW'), 
                   names_to = 'Model', values_to = 'SMD') %>% 
      mutate(flag = if_else(abs(SMD)>0.1, cbbPalette[1], cbbPalette[2])) %>%
      ggplot(., aes(y = Variable, x = SMD, group = Model)) +
      geom_point(aes(shape=Model, color = flag), size=1.8) +
      geom_vline(xintercept = 0, linetype=2) +
      geom_vline(xintercept = 0.2, linetype=2, color=cbbPalette[3]) +
      scale_color_identity() + 
      facet_grid(cols = vars(time)) +
      theme_classic() +
      labs(x = 'SMD', y = 'Variable',
           caption = 'M-dist = Mahalanobis distance for overall groups covariate means')
  
  gg_bal
```

::: callout-note
You can see the weighted between group differences for `X` and `X_t` are significantly reduced by the IPW. However, caution is advised in interpretation of the differences. If the treatment/exposure of interest has a causal relationship with the outcome (or some other censoring mechanism besides the artificial censoring the weights account for), then you might expect there to be differences between groups like `gender` and `age`. This is because the groups are conditional on remaining alive and uncensored at each time-point, which may occur after treatment/exposure. If other covariates also cause the outcome, you would expect distributions to diverge over time.
:::

## Reference papers in the literature  

# References

::: {#refs}
:::